% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{longtable,booktabs,array}
\newcounter{none} % for unnumbered tables
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{hyperxmp}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Most infrastructure layers are symptoms of the persistence model: a construct for auditing production stacks},
  pdfauthor={Alvaro Rivera},
  pdfkeywords={\xmpquote{infrastructural symptom}, \xmpquote{design
theory}, \xmpquote{persistence model}, \xmpquote{compensatory
layer}, \xmpquote{journaled systems}, \xmpquote{actor-native
architecture}, \xmpquote{Redis}, \xmpquote{DBMS-centric
architecture}, \xmpquote{appliance deployment}, \xmpquote{puppeteer
framework}, \xmpquote{accidental category}, \xmpquote{server-role}},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Most infrastructure layers are symptoms of the persistence model:
a construct for auditing production stacks}
\author{Alvaro Rivera}
\date{2026-05-15}

\begin{document}
\maketitle
\begin{abstract}
Production software systems routinely include layers such as caches,
object-relational mappers, message queues, distributed locks,
application server frameworks, and orchestration clusters. These layers
are treated as standard components of non-trivial deployments, yet they
are rarely examined under a common explanation for why they exist.

This paper argues that many of these layers do not arise from structural
requirements of the problem being solved, but from structural
deficiencies of the underlying persistence model. It names this property
\emph{infrastructural symptom}. A layer exhibits infrastructural symptom
when it compensates for a defect of the persistence model and loses
justification when that defect is removed. Two diagnostic indicators ---
compensation and dissolution --- give shape to this distinction, and an
auditor workflow guides their application to any (layer, use-case) pair
without requiring adoption of a specific alternative.

The paper presents a journaled actor-native system as an existence proof
in which the canonical symptoms dissolve. Four mechanisms --- locality,
journal density, journal-as-causal-substrate, and self-contained
substrate --- account for the disappearance of seven common categories
of compensating infrastructure. The construct extends through a ladder
of additional layers to a limit case in which whole categories of
business software become deliverable as a single appliance, as an
architectural permission rather than an operational demonstration.

The contribution is analytic: a diagnostic typology that makes visible
which parts of a stack are tied to the persistence model rather than to
the problem domain.
\end{abstract}

\section{Most infrastructure layers are symptoms of the persistence
model: a construct for auditing production
stacks}\label{most-infrastructure-layers-are-symptoms-of-the-persistence-model-a-construct-for-auditing-production-stacks}

\subsection{TL;DR}\label{tldr}

Many layers taken for granted in production systems --- caches, ORMs,
message queues, distributed locks, application servers, orchestration
clusters --- do not exist because the problem requires them, but because
the underlying persistence model does.

This paper names that property \emph{infrastructural symptom}, defines
the two diagnostic indicators that distinguish it from a structural
requirement, provides an auditor workflow to read any layer under those
indicators, and shows a journaled actor-native system in which the
canonical symptoms disappear.

The construct is diagnostic, not prescriptive: it makes visible which
parts of a stack compensate for the model rather than serve the problem.

\subsection{Claims this paper makes}\label{claims-this-paper-makes}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Some infrastructural layers in production deployments exist to
  compensate for structural deficiencies of the underlying persistence
  model rather than for structural requirements of the problem being
  solved (§3).
\item
  The property naming this distinction --- \emph{infrastructural
  symptom} --- is read through two diagnostic indicators: compensation
  (the layer compensates for an identifiable model defect) and
  dissolution (the layer loses justification when the defect is removed)
  (§3).
\item
  An auditor workflow guides the application of the two indicators and
  produces a reading --- model-dependent, structural, or unclassified
  --- per (layer, use-case) pair (§5).
\item
  A journaled actor-native model instantiates a system in which four
  mechanisms --- locality, journal density, journal-as-causal-substrate,
  self-contained substrate --- dissolve the seven canonical categories
  of compensating infrastructure (§4.1).
\item
  The Redis case admits the workflow with per-use-case reading: six
  use-cases dissolve as symptoms; two remain structural (§4.2).
\item
  The ladder of compensated layers is open-ended; the workflow extends
  to backup tooling, service mesh, distributed-tracing/APM, and future
  emergent layers (§6).
\item
  The model permits appliance-class deployment as a limit case in which
  categories of business software become physically shippable --- an
  architectural permission, not an operational proof (§7).
\end{enumerate}

\subsection{1. Introduction}\label{introduction}

This paper makes an analytic contribution: a diagnostic typology for
reading existing stacks. It identifies and names a structural pattern
that prior literature has documented case by case without articulating
as one (the typology); derives the indicators under which the pattern
reads as model-dependent (the criteria); and presents an instantiation
in which those indicators read positively and the pattern is absent (the
existence proof). The genre is closer to \emph{theory for analyzing} in
the sense of Gregor (2006) than to design theory: evidence is presented
in the form of a working artifact, but the artifact illustrates the
lens's reading rather than validating it. The contribution is
conceptual; the instantiation confirms that the pattern is contingent
rather than structurally necessary, not that any specific system is the
only way to dissolve it.

Production software systems are routinely deployed alongside layers of
supporting infrastructure: in-memory caches, object-relational mappers,
message queues, distributed locks, application server frameworks,
orchestration clusters, full operating systems. These layers are treated
as standard components of any non-trivial deployment. Architectural
discussions assume their presence by default; tutorials and reference
architectures position them as foundational; engineering teams budget
for them, staff them, and monitor them. Their absence is treated as a
property of toy systems.

This paper names a property that some --- not all --- of these layers
share. We call this property \emph{infrastructural symptom}: a layer is
an infrastructural symptom when it exists to compensate for a structural
deficiency in the underlying persistence model rather than for a
structural requirement of the problem being solved. The distinction is
routinely elided; the two cases are treated alike in architecture, in
tooling, and in engineering culture. Once the property is named, each
layer in a given stack becomes auditable under the question: is this
compensating for something, and if so, for what?

Section 2 situates the paper among prior work. Section 3 defines the
construct and derives the two diagnostic indicators that distinguish
infrastructural symptom from structural requirement. Section 4 presents
the instantiation: a journaled actor-native system in which the
canonical layers --- caches, glue queues, locks, application server
scaffolding --- are absent not by omission but by structural
irrelevance. Section 5 specifies the auditor workflow for reading a
stack under the typology. Section 6 generalizes from the canonical case
to a chain of compensated layers, from in-memory cache up through full
operating system. Section 7 examines the limit case, where the chain
collapses to a domain artifact deliverable as an appliance. Section 8
examines what the lens makes visible and what it does not.

\subsection{2. Related work}\label{related-work}

The genre of this paper is closer to \emph{theory for analyzing}
(Gregor, 2006) than to design theory. An analytic paper of this kind
identifies a typology, derives the indicators under which the typology
reads, and presents an instantiation that demonstrates the reading
produces non-trivial discrimination. The contribution is conceptual; the
instantiation is the existence proof, not the substance of the claim.

Critique of DBMS-centric architecture as the dominant paradigm for
production systems has accumulated in the literature without naming the
construct presented here. Stonebraker and Çetintemel (2005) observe that
the relational database has become a one-size-fits-all hammer applied to
use-cases for which it is structurally ill-suited, and predict the
proliferation of specialized stores. Kleppmann (2017) catalogs the
patterns by which contemporary deployments combine caches, queues, ORMs,
and search indices around a relational core, documenting their pairwise
interactions and operational costs. Helland's \emph{Life beyond
Distributed Transactions} (2007) makes a parallel observation about the
proliferation of coordination machinery under distributed-transaction
semantics. His \emph{Immutability Changes Everything} (2016) extends a
related argument to the substrate itself: append-only representation
dissolves much of that machinery by construction. The present paper
names what these critiques document piecewise --- the unifying property
of compensation --- and renders the discrimination available as a
diagnostic lens.

The instantiation presented in §4 is constructed from three lines of
prior work. The actor model originates with Hewitt (1973), articulating
computation as message passing between isolated entities. Event sourcing
as a persistence pattern is associated with Fowler (2005), which
characterizes the program's history as the durable artifact rather than
a snapshot of current state. Command Query Responsibility Segregation,
articulated by Young (2010), separates the read and write paths so they
no longer compete for the same model. Vernon (2015) brings these three
together as practical patterns in \emph{Reactive Messaging Patterns with
the Actor Model}; that synthesis is the immediate ancestor of the
present instantiation. The journaled actor-native system used as the
existence proof here combines these threads; it does not invent them.

This paper is the sixth in a series. The series develops the construct's
foundations across five preceding papers. Paper 1 introduces porosity as
the representational defect that this paper's construct identifies in
its compensation form. Paper 2 develops program-value separability and
the externalized-parameters precondition that supports journal density
as a §4.1 mechanism. Paper 3 introduces Reactions and the pragmatic
partition between work-due-before-responding and deferred work, which
§4.1 invokes as the journal-as-causal-substrate mechanism. Paper 4
introduces the Tell primitive and cross-actor causal continuity, used in
§4.1 for the same mechanism. Paper 5 develops the journal as substrate
for deployment, replication, backup, and offline operation, which §6
references as the substrate-mediated lifecycle that dissolves
application-server scaffolding. Paper 7 carries the same discrimination
upward from infrastructural layers to architectural roles; §9 introduces
this extension.

The contribution unique to this paper is conceptual. The construct
\emph{infrastructural symptom} names a property that prior work
documents in pieces but does not unify. The two diagnostic indicators
--- compensation and dissolution --- give the property its shape. The
auditor workflow of §5 guides its application without requiring adoption
of any particular alternative. The ladder of §6 and the limit case of §7
extend the typology's reach. The instantiation of §4 demonstrates that
the typology reads cleanly under the indicators; the demonstration is
not the contribution.

\subsection{3. The construct: infrastructural
symptom}\label{the-construct-infrastructural-symptom}

We state the construct before naming its two diagnostic indicators and
the workflow that guides their application.

An infrastructural symptom is a layer of infrastructure that exists to
compensate for a structural deficiency in the underlying persistence
model rather than for a structural requirement of the problem being
solved.

The construct operates at the level of purpose, not at the level of
product. A given physical layer in a deployment may serve multiple
purposes; the construct discriminates among those purposes one at a
time. The unit of analysis is the use-case, not the layer.

A layer reads as an infrastructural symptom for a given use-case when
both of two diagnostic indicators are positive.

\textbf{The compensation indicator.} The layer's purpose for that
use-case is traceable to a specific deficiency of the underlying
persistence model. The deficiency must be identifiable: not ``we use
this layer because it is fast,'' but ``we use this layer because the
persistence model exhibits property P, and this layer compensates for
P.'' When the compensation indicator is positive, asking what would
happen if P were not present? yields a non-trivial answer.

\textbf{The dissolution indicator.} When that deficiency is removed ---
by adopting a persistence model in which P does not arise --- the layer
loses justification for that use-case. The deficiency's absence is
sufficient to make the layer redundant for that purpose; nothing else
about the problem domain, workload, or operational context needs to
change.

The dissolution indicator requires that M' --- the counterfactual model
in which P does not arise --- be \emph{constructible}: a persistence
model with operationally defined semantics, viable under the workload,
hardware, and programming-model constraints of the use-case under
examination. A counterfactual whose viability is unspecified --- an
imagined absence of P without a model that exhibits the absence in
operational form --- does not count as a model in which D does not
arise; it only restates the wish that D were absent. The lens does not
admit such counterfactuals. The instantiation of §4 is offered precisely
as one such constructible M', sufficient to make the dissolution
indicator non-vacuous for the seven canonical categories.

Both indicators must be positive. Compensation without dissolution
describes a genuine patch whose value survives model evolution.
Dissolution without compensation is not constructible: a layer cannot
dissolve under a model change unless its purpose was tied to the model
in the first place.

The construct is not a universal accusation against infrastructure. Many
layers in production systems exist as structural requirements --- they
serve purposes rooted in the problem domain, independent of any
persistence model. Three categories are typical:

\begin{itemize}
\tightlist
\item
  \textbf{External-system integration.} Layers mediating communication
  with systems outside the boundary --- payment gateways, partner APIs,
  sensor networks --- compensate for no model deficiency. The
  integration problem remains under any persistence model.
\item
  \textbf{Cryptographic and protocol enforcement.} TLS termination,
  request signing, OAuth verification address security or protocol
  requirements rooted in the problem domain. No persistence-model change
  makes encryption optional.
\item
  \textbf{Genuinely expensive computation.} Analytics engines,
  machine-learning inference services, video encoders exist because the
  computation itself is intrinsically expensive. That cost does not
  change under a different persistence model.
\end{itemize}

The boundary between symptom and structural requirement is not a
property of the layer as a product but of the purpose for which it is
used. The same caching system, deployed to absorb read-vs-write
contention at the persistence layer, instantiates an infrastructural
symptom. Deployed instead to memoize a deliberately expensive
computation, it instantiates a structural requirement. The reading is
produced per use-case.

Seven categories of layer are encountered regularly in production
deployments and admit reading under the two indicators. The third column
describes, in model-property terms rather than system terms, what a
persistence model under which both indicators read positively would
offer instead.

These seven are not a privileged set but an entrenched one: the layers
most consistently assumed by default in conventional production stacks,
taught as foundational, and budgeted as standing operational concerns.
Other commonly-deployed layers --- logging frameworks, schedulers,
service discovery, secret managers, feature-flag servers --- admit
reading under the same workflow of §5 and produce more varied results.
Some dissolve (in-process logging, internal scheduling and discovery);
some are structural (secret managers, whose purpose is rooted in access
control rather than persistence; the A/B-testing facet of feature-flag
servers, rooted in product methodology); some split per use-case
(logging at internal vs.~external boundaries; service discovery at
internal vs.~external). The lens discriminates; Table 1 catalogs the
architectural baseline against which the discrimination is most
consequential.

\textbf{Table 1.} Diagnostic reading: seven canonical categories of
compensating infrastructure under the two indicators of §3.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Layer
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Underlying defect compensated
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
What a model with both indicators positive offers instead
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
In-memory cache & Reads contend with writes at the persistence layer;
reconstructing state from storage is expensive & State local to the unit
of computation; no contention between read and write paths \\
Object-relational mapper & Domain objects do not align with relational
rows; mapping is opaque and lossy & Domain objects are the unit of
persistence; no translation layer \\
Message queue used as glue & Components cannot directly invoke deferred
work without losing causal record & Deferred work is journal-recorded
and replayable; queues are not the substrate of causality \\
Distributed lock & Multiple writers contend for the same logical entity
across processes & A single point of authority per logical entity;
serialization is intrinsic, not externalized \\
Application server framework & The persistence model leaks into
application code; lifecycle plumbing must be added & The domain is the
application; lifecycle is a property of the substrate, not added
scaffolding \\
Orchestration cluster as coordination substrate & Stateful coordination
across instances requires external choreography & State has location;
coordination is between actors, not between containers \\
Full operating system & The runtime, the framework, and the application
require disparate substrates & A minimal substrate suffices to run the
domain artifact \\
\end{longtable}
}

The seven rows are not exhaustive. Backup tooling, service mesh, and
distributed-tracing frameworks admit similar analysis and are discussed
in Section 6. Whether such a persistence model exists, and at what cost,
is the subject of Section 4.

The construct is diagnostic, not prescriptive. It enables auditing
existing stacks under the two indicators; it does not specify how to
build a model in which infrastructural symptoms do not arise.
Construction is the work of Section 4, which exhibits one such model,
and of Section 5, which specifies the auditor workflow that produced the
diagnostic table.

The value of the typology is diagnostic. A practitioner who does not
adopt the instantiation of Section 4 can nevertheless use it to read
which layers in their stack are tied to their persistence model and
which are tied to their problem domain. That visibility is independent
of the choice to act on it.

\subsection{4. Instantiation}\label{instantiation}

\subsubsection{4.0 Genealogy}\label{genealogy}

Puppeteer's earliest layer dates to 2005, when a persistence library
internally called \emph{autopersistencia} adopted a single design rule:
the classes representing the domain would carry no persistence concerns.
Persistence would be discovered by reflection over the class hierarchy;
storage details lived elsewhere. The rule was modest in scope and
pragmatic in motivation. Its consequence was structural: the domain
became a clean substrate, unobstructed by the technical aspects that
conventional architectures embed inside it.

At no point in this evolution was there an intention to eliminate
caches, queues, locks, or application scaffolding; those layers were
assumed to belong in any serious system. Their later absence was not a
design objective but an empirical outcome. This distinction matters for
the argument of this paper: the construct is not dissolved here by
architectural ideology but as a byproduct of unrelated design
constraints that happened to satisfy the indicators derived in §3.

The natural follow-up question was: if domain classes carry no
persistence concerns, where does state live? The answer that emerged
across the following years became the framework's first generation.
State lives in a journal, and the journal records the program itself ---
not serialized events, not row deltas, but the DSL script that produced
the state, replayable in order. The narrative of execution, not the
photograph of memory, became the unit of persistence --- a property
whose absence, in the terms of §3, is the deficiency that caches, ORMs,
and queue-glue exist to compensate for in DBMS-centric architectures.

The framework's second generation added four properties, each developed
in preceding papers of this series. Parameters were externalized,
allowing programs to be compiled, cached, and recorded as references
rather than literal scripts (Paper 2). Reactions were introduced as the
developer-controlled partition between work due before responding and
work deferred to placement (Paper 3). The Tell primitive established
cross-actor causality recorded in the journal rather than orchestrated
externally (Paper 4). The journal itself became the catalog of the
actor's declarative vocabulary, dissolving the lateral storage of action
definitions that the framework had carried since V1.

The framework was not designed to dissolve the construct named in this
paper. The construct was identified retrospectively, after the chain of
architectural decisions described above had converged on a system in
which most of the canonical layers had become unnecessary. The work of
this paper is, in that sense, a forensic reading of an existing
artifact, not an engineering plan derived from a principle. The
principle was a property of the artifact before it was a property of the
literature.

\subsubsection{4.1 The instantiation under the two
indicators}\label{the-instantiation-under-the-two-indicators}

Four mechanisms in the journaled actor-native model account for the
dissolution of the seven canonical categories of §3. A single mechanism
may dissolve more than one category, and the in-memory cache dissolves
under two mechanisms because §3 records two distinct defects compensated
by that layer. Each mechanism decomposes into several sub-mechanisms ---
properties of the model that, taken together, exhibit the umbrella
behavior. Table 2 correlates each mechanism with its sub-mechanisms, the
§3 defects it dissolves, and the canonical layers that become symptoms
under it.

\textbf{Table 2.} Mechanisms in the journaled actor-native model mapped
to the §3 defects they dissolve and the canonical layers that become
symptoms.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Mechanism
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
§3 defect compensated
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Canonical layer that becomes symptom
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Locality}~~• privacy of state~~• actor-as-serializer~~•
placement & Reads contend with writes at the persistence layerMultiple
writers contend for the same logical entity & In-memory cache
(contention defect)Distributed lock \\
\textbf{Journal density}~~• journal-as-program (homoiconic)~~•
replay-as-reconstruction~~• domain-class-as-unit-of-persistence &
Reconstructing state from storage is expensiveDomain objects do not
align with relational rows & In-memory cache (reconstruction
defect)Object-relational mapper \\
\textbf{Journal as causal substrate}~~• Reactions recorded in journal~~•
Tell recorded in journal~~• causality boundary is the journal &
Components cannot directly invoke deferred work without losing causal
recordStateful coordination across instances requires external
choreography & Message queue used as glueOrchestration cluster \\
\textbf{Self-contained substrate}~~• domain artifact is the
application~~• substrate carries lifecycle, persistence, dispatch~~•
journal-mediated release handoff~~• minimal OS surface & The persistence
model leaks into application codeThe runtime, framework, and application
require disparate substrates & Application server frameworkFull
operating system \\
\end{longtable}
}

\textbf{Locality.} State is private to the actor that owns it; the actor
processes commands and queries serially against its own memory. There is
no shared store from which concurrent reads must be isolated from
concurrent writes. Two of the §3 defects vanish mechanically under this
property: the read-versus-write contention that motivates in-memory
caches, and the inter-process contention that motivates distributed
locks. The compensation indicator is vacuous for both. Placement is the
developer-controlled partition treated in Paper 3. This mechanism reads
positively under the actor-native preconditions established across the
preceding papers in this series: the actor's state fits in memory and is
processed serially over its private storage (Papers 1 and 3), and
cross-actor continuity is mediated through the journal by Reactions and
Tell rather than through shared queries (Papers 3 and 4). Aggregates
that exceed the in-memory precondition --- extremely hot single
entities, datasets larger than a single host's working set --- fall
outside this reading and require actor splitting or partitioning, which
is a design technique distinct from the lens itself.

\textbf{Journal density.} The journal does not record serialized
snapshots or row-by-row deltas; it records the program --- the DSL
script --- that produced the state. Reconstruction is replay, not
deserialization; replay of a compact script is cheap. The domain class
is the unit of persistence, with no intermediate row to map. Both §3
defects compensated by caches and ORMs --- reconstruction expense and
object-row impedance --- are dissolved mechanically. The mechanism is
treated as compilation in Paper 2 and as homoiconic representation in
Paper 1. Implementation: \texttt{Puppeteer/EventSourcing/DB/Diary.cs}.
Replay cost for long-lived actors is bounded by complementary substrate
primitives: Distill compacts the journal in place
(\texttt{Puppeteer/EventSourcing/DB/Diary.cs:264}); elision markers and
Materialize watermarks enable offload-and-trim through a materialization
destination (Paper 5, §7.3). Together these primitives fill the role
that Rolling Snapshots (Young, 2010) play in conventional event-sourced
systems, without re-introducing snapshot semantics.

\textbf{Journal as causal substrate.} Pragmatic deferral, developed in
Paper 3 as Reactions, records work the verb does not perform before
responding as a journal entry, with the causal link to the originating
event preserved. Cross-actor continuity, developed in Paper 4 as Tell,
records dispatch from one actor to another in the journal as well. The
journal becomes the boundary of causality. In DBMS-centric systems,
causality crosses the persistence boundary and must be reconstructed
with message queues used as glue and orchestration clusters used as
coordination substrates; here, causality never leaves the journal.
Implementation: the \texttt{reactions} field in
\texttt{Puppeteer/EventSourcing/ActorHandler.cs:38}.

\textbf{Self-contained substrate.} The runtime is a small loader; the
application is the domain artifact. Lifecycle, persistence, and dispatch
are properties of the substrate, so no external process is required to
host or coordinate the application artifact. Release handoff and rolling
deployment, handled in DBMS-centric deployments by application server
orchestration, are journal-mediated here and are the subject of Paper 5.
The application server framework as a category has no equivalent role;
the operating system surface is reduced to the minimum required, a limit
case treated in §7.

Together, these four mechanisms account for the dissolution of the seven
canonical categories of §3 mechanically, not by intent. None was added
to the model with the construct of this paper in mind; each was
developed for the reasons enumerated in §4.0.

\subsubsection{4.2 The Redis case in
detail}\label{the-redis-case-in-detail}

Redis occupies a distinctive position in the contemporary infrastructure
landscape: a single product whose presence is taken for granted across
deployments that are otherwise architecturally dissimilar. The diversity
of its typical use-cases --- cache, session store, pub/sub, queue,
distributed lock, leaderboard, rate limiter --- does not reflect the
diversity of a single problem domain. It reflects the diversity of the
pressures that DBMS-centric architectures present to application
designers, each of which Redis conveniently absorbs in a familiar form.
This makes Redis the canonical case for inspection under the construct
of §3. The analysis that follows is not, however, about Redis as a
product; it is about the defects of the underlying persistence model
that Redis exists to absorb --- defects that other compensating layers
also address in variant forms, treated briefly after the canonical
table.

Table 3 applies the two indicators of §3 to the most common Redis
use-cases, tracing each to the specific defect it compensates and the
mechanism in §4.1 that dissolves it.

\textbf{Table 3.} Application of the diagnostic typology to canonical
Redis use-cases.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Redis use-case
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
§3 defect compensated
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Native response in journaled actor-native
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
In-process cache for database results & Reads contend with writes at the
persistence layer; reconstructing state from storage is expensive &
Actor state is the cache (§4.1 Locality + Journal density) \\
Session storage & Stateless application servers cannot retain user state
across requests & The user actor retains session naturally as part of
its state (§4.1 Locality + Self-contained substrate) \\
Pub/sub among internal components & No causal record of cross-component
events; coordination must be reconstructed & Reactions record deferred
work in the journal with causal links preserved (§4.1 Journal as causal
substrate) \\
Message queue for deferred work & Components cannot directly invoke
deferred work without losing causal record & Reactions and Tell record
causation in the journal (§4.1 Journal as causal substrate) \\
Distributed lock for shared entity & Multiple processes write the same
logical entity across boundaries & The actor owns its entity;
serialization is intrinsic, not externalized (§4.1 Locality) \\
Leaderboard / sorted set & Real-time aggregation against the persistence
layer is expensive & Query against actor state or a follower; replay is
cheap (§4.1 Locality + Journal density) \\
\end{longtable}
}

Each row reads positively under both the compensation and dissolution
indicators of §3. The pub/sub row covers internal-component
coordination; the external-systems variant is structural and is treated
below.

Two Redis use-cases do not read positively under the dissolution
indicator and are therefore structural under the typology. The first is
pub/sub directed at external systems: webhooks to third-party consumers,
event distribution to partners, broadcast to subscribers outside the
operator's boundary. The external system does not dissolve under any
persistence-model change; the distribution problem persists. The second
is rate limiting applied to external API consumption: per-customer
quotas, third-party API throttling, abuse mitigation against external
traffic. The rate limit is a property of the external relationship, not
of the internal model. Redis, or any comparable substrate, remains the
appropriate tool for both.

Other compensating layers in the DBMS-centric ecosystem exhibit variants
of the same pattern. Memcached shares the cache analysis: locality and
journal density make in-process caches mechanically redundant.
Distributed coordinators such as ZooKeeper share the lock and
coordination analysis: the actor model makes the underlying contention
vacuous. Message brokers such as RabbitMQ share the queue analysis:
Reactions and Tell record causation in the journal.

Event-storage products such as EventStore DB and Kafka deserve a
separate observation. Both adopt a journaled posture --- they record
events rather than serialized snapshots, and they offer replay. But they
instantiate the journal as an external infrastructure layer that the
application consumes through a client API, not as the substrate of the
application itself. Causality is recorded but remains external to the
unit of computation; coordination between event production, event
consumption, and application state still requires application-side glue.
The product captures the right idea and externalizes it, leaving the
application code in the compensating position. Under the construct,
these products do not dissolve the symptom; they relocate it.

A second observation about these compensating layers is that they do not
absorb their compensation freely. Each requires the application to carry
the boilerplate of its use: cache invalidation logic, lock acquisition
and release with retry and fencing semantics, serialization between
domain objects and external representations, key naming and TTL
discipline, consumer offset management and rebalancing handlers, schema
registry coordination, idempotency keys for at-least-once delivery. The
application code carries the porosity imposed by RDBMS representations
(Paper 1) as its baseline; the compensating layer adds a second
porosity, this time imposed by the layer itself: its own API surface,
lifecycle, and consistency contracts. Domain logic interleaves with
both. Under the instantiation, the compensating layer and the
boilerplate that the application carries to use it are both absent. The
actor's code addresses the problem domain; the substrate addresses the
rest.

Redis as a product is not what the typology identifies; nor is any
single compensating layer treated above. What the typology reads is the
architectural pattern in which a persistence model exposes defects and
successive layers --- caches, brokers, coordinators, even externalized
event stores --- accumulate to absorb them. Replacing one such layer
while retaining the model that produces the others does not change how
the typology reads. The unit of change is the persistence model itself,
not any individual layer.

\subsection{5. Auditor workflow}\label{auditor-workflow}

The typology of §3 enables reading per use-case, but practitioners face
stacks containing dozens of compensating candidates. Without an explicit
workflow for application, the typology risks remaining theoretical:
invoked at the level of architecture review, but absent from the
operating discipline that distinguishes one layer from the next. The
workflow that follows guides the application of the two indicators of
§3, producing a reading in a finite number of steps for any (layer,
use-case) pair under examination. It can be applied without adopting the
instantiation of §4.

\begin{verbatim}
Input:  layer L deployed for use-case U.
Output: reading in {model-dependent, structural, unclassified};
        defect D identified when model-dependent.

Step 1 — Compensation indicator
  Identify the defect D of the underlying persistence model
  for which L compensates in use-case U.
  If a defect D is identifiable, proceed to Step 2.
  If no defect is identifiable AND L's purpose is rooted in the
    problem domain (external-system integration, cryptographic
    or protocol enforcement, intrinsically expensive computation),
    return structural.
  If no defect is identifiable AND L's purpose is opaque to the
    auditor (organizational, regulatory, historical, team-skill),
    return unclassified.

Step 2 — Dissolution indicator
  Consider a counterfactual model M' in which D does not arise.
  M' must be constructible: a persistence model with operationally
    defined semantics, viable under the workload, hardware, and
    programming-model constraints of U.
  If no such M' is constructible, return unclassified.
  If L retains a role in M' for U, return structural.
  Otherwise return model-dependent with D.
\end{verbatim}

Four notes on application. First, the unit of analysis is the pair
(layer, use-case), not the layer alone. The same product may produce
different readings for different uses; the workflow must be run once per
use-case. Second, identifying the defect in Step 1 requires explicit
attribution to a property of the persistence model. \emph{``We use this
layer because it is faster''} does not name a defect; \emph{``we use
this layer because the model serializes writes against a shared store''}
does. The workflow depends on the discipline of this attribution. Third,
the \emph{unclassified} reading exists precisely so that the workflow
does not silently absorb opaque or non-model-rooted layers as structural
by default; an honest auditor records what the lens cannot see rather
than forcing a binary outcome. Fourth, the diagnostic table of §3 and
the Redis case of §4.2 are worked examples of the workflow applied to
canonical and product-specific layers respectively. Each row in those
tables is the reading produced by the workflow for one (layer, use-case)
pair.

The workflow renders readings; it does not prescribe action. A
model-dependent reading does not require that the layer be removed in
the current system. It indicates that the layer's continued presence is
structurally tied to retaining the persistence model that produces the
defect. Practitioners may rationally retain a model-dependent layer ---
for legacy preservation, contractual constraint, team familiarity,
migration cost --- but they do so knowing that the layer carries a debt
that another model would not impose. The workflow's product is
visibility, not directive.

\subsection{6. The ladder of compensated
layers}\label{the-ladder-of-compensated-layers}

The seven canonical categories enumerated in §3 are not a list. Read in
order, they form a ladder of increasing scale: in-memory cache
(per-process), object-relational mapper (within-application), message
queue and distributed lock (between processes), application server
framework (the application stack), orchestration cluster (across
machines), and full operating system (the hosting substrate itself). The
order is not arbitrary; it follows the expanding radius of the defect
the layer compensates for. Each rung compensates for a deeper limitation
of the model; each rung dissolves under the alternative for the same
reason.

Three additional categories that §3 forward-referenced extend the
ladder.

Backup and disaster-recovery tooling --- point-in-time restore,
write-ahead log archiving, snapshot orchestration --- exists because
state in DBMS-centric architectures lives in mutable rows whose history
is not the primary artifact. Under the model, the journal is the backup,
and rehydration is the recovery operation; both reduce to a single
substrate property treated more fully in Paper 5.

The service mesh --- inter-service routing, discovery, traffic shaping,
mTLS, sidecar instrumentation --- exists because stateless application
servers, by design, do not retain coordination state internally. Under
the model, much of this coordination is internal: Reactions and Tell
handle deferred work and cross-actor dispatch as journal entries. The
parts that cross the operator's boundary (TLS to external partners,
traffic shaping for external clients) remain structural and continue to
belong in a mesh or comparable substrate.

The distributed-tracing and APM stack --- span propagation, trace
ingestion, request-flow visualization, latency aggregation --- requires
discrimination per use-case, treated in Table 4 below.

\textbf{Table 4.} Use-case-level reading for the distributed-tracing and
APM stack.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Observability use-case
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Reading
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Reason
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Distributed tracing to reconstruct what happened inside the system &
Model-dependent & The journal already records the trace; replay is
deterministic \\
Distributed tracing across boundaries with external systems & Structural
& The external system does not dissolve under any persistence-model
change \\
APM as \emph{``find the bug in this incident''} & Model-dependent
(predominantly) & Replay localizes the bug without requiring external
instrumentation \\
APM as continuous capacity, latency, and error-rate telemetry in
production & Structural & Resource consumption is a real-world property,
not a model artifact \\
\end{longtable}
}

The first and third rows read as model-dependent; the second and fourth
read as structural requirements. Debug as a one-shot operation dissolves
more cleanly than continuous telemetry because the journal substitutes
directly for the debugger. Continuous telemetry is mixed: the
internal-trace component dissolves, but the capacity-and-cross-system
component remains. §8 treats this as a worked example of the typology's
reading at the boundary of its applicability.

The typology's readings do not eliminate the products from a deployment.
The framework does not dissolve APM and OTel; it dissolves the
internal-tracing use-case while leaving APM and OTel plugged in at the
points where they instantiate a structural requirement. The same
observation applies to mesh products at external boundaries, and to
backup tooling for legacy systems retained alongside a journaled core.

The ladder is not closed. New layers will emerge to compensate for new
defects of DBMS-centric architectures --- feature-flag servers, schema
registries, secret stores, sidecars for auth and identity,
configuration-as-a-service products. Each new layer can be audited under
the workflow of §5 and read under the typology of §3. The ladder is open
upward, sideways, and into the future because the underlying persistence
model continues to generate compensations; the workflow walks each new
rung as it appears.

\subsection{7. At the limit: from datacenter footprint to a deliverable
appliance}\label{at-the-limit-from-datacenter-footprint-to-a-deliverable-appliance}

The workflow of §5 walks each rung of the ladder as it appears; §6
showed the ladder extends. The natural question is what remains when the
workflow of §5 marks every rung as model-dependent. §7 names that limit.

What remains is the domain artifact, the journal that records its
execution, and the minimal substrate that runs them. The substrate
retains what is irreducibly needed: a runtime, journal storage, network
I/O, and a scheduling loop. Not required are the cluster orchestrator,
the application server framework, the object-relational mapper, the
external cache, the distributed coordinator, and the general-purpose
operating system in its hosting form. The domain artifact is the
application; the substrate is what is required to run it; everything
else has been audited and removed under the typology.

The artifact's architectural requirements are independent of scale. The
same architecture sits behind a multi-datacenter cluster with
cross-region replication (where the journal and Tell handle topology),
behind a Kubernetes red-black rolling deployment (where the journal
mediates the release handoff), behind a single server (where the
artifact runs against local storage), and behind an on-premises
appliance hosting an accounting system, a clinic management system, a
hardware-store inventory and point-of-sale system, or a
professional-services back office. The substrate's storage requirement
reduces to two operations --- appending entries and advancing a cursor
over them --- which together compose the substrate's single primitive of
replay (Paper 5); both are executed by commodity flash at marginal cost.
The RAM requirement scales with the domain's working state, not with the
deployment topology. The difference between these shapes is operational
variation over the same architecture. What bounds the appliance case is
the customer's domain --- its data volume, user count, workload
heaviness --- not anything the architecture introduces.

This is an engineering-compatibility claim, not an operational
demonstration. The deployments of the framework's home site run in
conventional datacenter infrastructure across eight domain
installations; no appliance-class on-premises shipment is currently
among them. What §7 argues is that the path to that shape is bounded by
ordinary engineering parameters --- runtime footprint, journal latency
on flash storage, fleet management for updates --- and not by an
architectural obstacle that the lens has not already named. The barriers
that conventional architectures encounter at the appliance limit (a DBMS
that does not fit the box and its administration overhead, an
application server that demands a separate tier, an orchestrator with no
peer to coordinate with, a cache for a model that does not need it) are
precisely the layers the workflow of §5 reads as symptoms. The model's
permission for the appliance shape is the structural counterpart of
those symptoms' dissolution.

This bears restating because of what it implies. Whole categories of
small-and-medium-business software --- accounting, clinic management,
retail inventory, professional services --- are today delivered on
infrastructure scaled to a fraction of their actual workload, because
the conventional stack obliges a datacenter footprint for a database, an
application server, and the orchestration to keep them running. The
barrier is not only to delivery but to development: building software
for a single hardware store or a single clinic requires the
infrastructural expertise of running a datacenter, which is the
structural reason such products often do not exist at all. The lens
reads each of those layers as a symptom. What §7 names is the limit case
made available by that reading: software that is shippable as hardware
to the customer, hosted in a single appliance, operated by the customer
or a local service provider.

At the structural limit, the architecture permits software to be
delivered as a physical artifact: an appliance the customer owns, not a
datacenter the vendor leases on their behalf. The barrier to that shape
is not in the application or in the customer's needs --- it is in the
infrastructure that conventional architectures require. The barrier
dissolves with the symptoms.

\subsection{8. What the lens makes visible and what it does
not}\label{what-the-lens-makes-visible-and-what-it-does-not}

Sections 3 through 7 showed the typology in action: canonical layers in
§3, the Redis case in §4, the workflow in §5, the extended ladder in §6,
and the limit case in §7. Each of those sections assumed conditions ---
that the persistence model is identifiable, that use-cases are
decomposable, that a counterfactual model is constructible. §8 treats
what the lens makes visible and what it does not. The typology still
applies; the readings become coarser and the translation to action grows
more complicated.

The canonical boundary case is observability. Table 4 (§6) shows that
the APM stack contains four use-cases with mixed readings: two
model-dependent (internal distributed tracing, one-shot debug) and two
structural (cross-system tracing, continuous capacity telemetry). Each
reading is correctly assigned per use-case. But a single OTel deployment
in a production system serves all four use-cases simultaneously without
surgical separation. Removing the instrumentation that serves the
internal-tracing use-case affects the instrumentation that serves
cross-system tracing; the two share span propagation, collector
pipelines, and storage backends. The typology reads the use-case, not
the deployment. Operational reality does not admit the same separation,
and the practitioner retains the deployment intact and carries the
model-dependent use-cases with it.

Four boundaries describe what the lens makes visible and what it does
not.

\textbf{Quantification.} The workflow returns a reading per use-case; it
does not predict cost saved, performance gained, or operational
complexity reduced. Quantification is the work of instrumentation, not
of a diagnostic typology --- distinct from a limitation, this is a
property of the genre.

\textbf{Stack prescription.} A stack populated by model-dependent layers
can be identified as suboptimal, but the typology does not prescribe an
alternative. The instantiation of §4 demonstrates that one exists; it
does not specify the unique alternative for any given system.

\textbf{Adoption prediction.} A model-dependent reading does not inform
the difficulty of migration: organizational costs, contractual
constraints, vendor lock-in, and engineering skills carry independent
weight.

\textbf{Opaque models.} The workflow requires identifying the defect of
the underlying persistence model in Step 1 of §5. Systems whose
persistence model is opaque --- third-party SaaS exposing only an API,
proprietary platforms without published internals --- fall outside its
applicability and are recorded as \emph{unclassified} under the
workflow's third reading.

Section 5's closing observation extends here: the visibility that the
lens renders is independent of the action the practitioner takes.
Model-dependent layers may remain in deployments for legitimate reasons
--- legacy preservation, contractual obligation, organizational
constraint, migration cost. The typology documents the debt; it does not
resolve it.

The lens reads cleanly where the persistence model is explicit and
characterizable, where use-cases are decomposable, and where a
counterfactual model is constructible. Where one of these conditions
fails, the typology still applies but its readings grow coarser, or the
workflow falls back to \emph{unclassified}. Where it reads cleanly, what
the lens surfaces is not a smaller stack but a domain that the model
permits to be modeled, libraried, and evolved without being interleaved
with compensating concerns. The utility of the typology is proportional
to the decomposability of the system under examination.

Persistence is one axis along which representational pressure deforms a
domain --- the axis this paper has traced. There is at least one more.
When interaction is modeled before the domain --- when aggregates take
the shape of the screen rather than the screen rendering the actor's
output --- the same compensation pattern reappears in a different
register: UI components, view-models, request DTOs, and step-state
objects accumulate around a domain that no longer fits its own
substrate.

The lens of accidental category reads there with the same discrimination
it produced for infrastructural layers and for the server-role. The
actor's existing output boundary --- the primitives by which it emits to
its invoker, agnostic of who consumes --- already factors transport out
of the domain. This axis is not developed here.

\subsection{9. Extension: from layers to
roles}\label{extension-from-layers-to-roles}

The typology of §3 reads among \emph{layers} of infrastructure ---
components that occupy positions in a stack and compensate for the
persistence model from those positions. Architectural \emph{roles} ---
the entity that accepts authoritative writes, the master in replication,
the bootstrap issuer, the orchestrator --- admit the same reading under
the same two indicators. A role is an accidental category when its
necessity is traceable to a specific representational choice (what nodes
replicate to share state) and the role loses justification when that
choice is replaced.

The same lens therefore applies beyond layers. Under a substrate in
which nodes replicate programs rather than data, state, or operations,
the server-role does not survive as a structural requirement. Its
dissolution is not an architectural objective; it is a structural
consequence of the same representational choice that dissolves the
layers described in §6.

The two analyses operate at different levels of abstraction --- first on
layers, then on the roles that those layers exist to serve --- without
modifying the typology itself. The reading scales upward unchanged.

\subsection{Code provenance}\label{code-provenance}

Source-code references in this paper resolve against the public
Puppeteer repository at commit
\href{https://github.com/alvaroNCubo/puppeteer/tree/2f31f9674a5de816bdf1bf9d8360ff218a02e4da}{\texttt{2f31f96}}
(2026-05-18). The snapshot is archived in Software Heritage under the
following persistent identifier:

\begin{verbatim}
swh:1:dir:10e7e6bad7eb77b6c2e406762026177f95c3ae92;
  origin=https://github.com/alvaroNCubo/puppeteer;
  anchor=swh:1:rev:2f31f9674a5de816bdf1bf9d8360ff218a02e4da
\end{verbatim}

Inline references of the form \texttt{file.cs:NN} (e.g.,
\texttt{ActorHandler.cs:38}) resolve against this snapshot. A reader can
construct a per-file SWHID by adding the qualifiers
\texttt{;path=\textless{}path\textgreater{};lines=\textless{}NN\textgreater{}}
to the directory SWHID above. Future commits to the repository may
renumber lines; the SWHID preserves the cited state independently of any
future change to the repository or its hosting.

\subsection{Acknowledgments}\label{acknowledgments}

The author used large language models (including Claude and ChatGPT) as
editorial assistants for language refinement, structural feedback, and
literature navigation. All original ideas, terminology, theoretical
constructs, and technical content presented in this work are solely the
author's.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{References}\label{references}

Fowler, M. (2005). Event sourcing. Retrieved from
https://martinfowler.com/eaaDev/EventSourcing.html

Gregor, S. (2006). The nature of theory in information systems.
\emph{MIS Quarterly}, 30(3), 611--642.

Helland, P. (2007). Life beyond distributed transactions: An apostate's
opinion. \emph{Proceedings of the 3rd Biennial Conference on Innovative
Data Systems Research (CIDR)}, 132--141.

Helland, P. (2016). Immutability changes everything.
\emph{Communications of the ACM}, 59(1), 64--70.

Hewitt, C., Bishop, P., \& Steiger, R. (1973). A universal modular ACTOR
formalism for artificial intelligence. \emph{Proceedings of the 3rd
International Joint Conference on Artificial Intelligence (IJCAI)},
235--245.

Kleppmann, M. (2017). \emph{Designing data-intensive applications: The
big ideas behind reliable, scalable, and maintainable systems}. O'Reilly
Media.

Rivera, A. (2026a). Anti-porous architecture: a unified design principle
for CQRS + Actor + Event-Sourcing systems. \emph{Puppeteer Papers
Series}, Paper 1.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/01-anti-porosity.md

Rivera, A. (2026b). Program--value separability: the structural
precondition for compilation, caching, and dense journaling in a DSL
runtime. \emph{Puppeteer Papers Series}, Paper 2.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/02-program-value-separability.md

Rivera, A. (2026c). Reactions and the partition: opt-in eventual
consistency in actor-native systems. \emph{Puppeteer Papers Series},
Paper 3.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/03-reactions-and-partition.md

Rivera, A. (2026d). Preserving semantic continuity across actors: a
tell-based approach without orchestration. \emph{Puppeteer Papers
Series}, Paper 4.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/04-cross-actor-continuity.md

Rivera, A. (2026e). The journal as substrate: unifying deployment,
replication, backup, and offline operation in distributed systems.
\emph{Puppeteer Papers Series}, Paper 5.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/05-substrate-operations.md

Rivera, A. (2026g). The server is not a structural requirement:
identifying accidental architectural roles under journaled programs.
\emph{Puppeteer Papers Series}, Paper 7.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/07-server-as-accidental-category.md

Stonebraker, M., \& Çetintemel, U. (2005). ``One size fits all'': An
idea whose time has come and gone. \emph{Proceedings of the 21st
International Conference on Data Engineering (ICDE)}, 2--11.

Vernon, V. (2015). \emph{Reactive messaging patterns with the actor
model: Applications and integration in Scala and Akka}. Addison-Wesley.

Young, G. (2010). CQRS documents. Retrieved from
https://cqrs.files.wordpress.com/2010/11/cqrs\_documents.pdf

\end{document}
