% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{longtable,booktabs,array}
\newcounter{none} % for unnumbered tables
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{hyperxmp}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The server is not a structural requirement: identifying accidental architectural roles under journaled programs},
  pdfauthor={Alvaro Rivera},
  pdfkeywords={\xmpquote{server-role}, \xmpquote{journaled
programs}, \xmpquote{distributed systems}, \xmpquote{design
theory}, \xmpquote{event sourcing}, \xmpquote{actor
model}, \xmpquote{bootstrap}, \xmpquote{local-first}},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{The server is not a structural requirement: identifying
accidental architectural roles under journaled programs}
\author{Alvaro Rivera}
\date{2026-05-16}

\begin{document}
\maketitle

\subsection{TL;DR}\label{tldr}

The role we call \emph{server} --- the entity that owns coordination,
accepts authoritative writes, and mediates between clients --- is
treated across the architecture literature as a structural component of
any non-trivial distributed system. Prior work has questioned the
server's privileges (local-first, federation), eliminated it as a design
goal (Hypercore, Holochain), or replaced it with a different topology
(peer-to-peer, blockchain). None of these positions removes the
server-role as a \emph{prerequisite for software construction itself}.

This paper names the property that distinguishes the server-role from
structural requirements of distributed software: it is an
\emph{accidental category}. Under a class of systems in which nodes
replicate programs rather than data or state, the server-role does not
survive as a necessity. The bootstrap of a new node, the last role that
appeared to require a running service, can be performed by an analog
artifact: a printed QR code transferred by any means and authorized by a
human decision.

The construct is recognition, not prescription. It enables auditing
distributed systems for whether the server-role exists in them as a
structural requirement or as a contingent artifact of the persistence
model. An instantiation --- a port of the Ordering bounded context of
\texttt{dotnet/eShop} to a journaled-program substrate, replicated
across three nodes that bootstrap through paper QR codes --- confirms
that the predicate dissolves under the conditions.

The contribution is conceptual; the instantiation is the existence
proof, not the substance.

\subsection{Formal statement}\label{formal-statement}

\textbf{Definition.} A \emph{server-role} is the architectural function
of accepting authoritative writes, owning coordination, mediating
between clients, and providing the address against which other parties
direct their traffic in a distributed system. The role is independent of
any specific machine, deployment topology, or cloud provider; it is the
function, not its incarnation.

A server-role is an \emph{accidental category} in a class of distributed
systems S when both of two conditions hold:

\begin{itemize}
\tightlist
\item
  \textbf{Contingency.} The role's necessity in S is traceable to a
  specific representational choice --- what the nodes of S replicate in
  order to share state.
\item
  \textbf{Dissolution.} The role loses justification when that
  representational choice is replaced; nothing else about the problem
  domain, workload, or operational context needs to change for the role
  to become unnecessary.
\end{itemize}

\textbf{Claim.} Under a class of systems in which nodes replicate
programs rather than data, state, or operations, both conditions hold
for the server-role. The conditions are stated in §3; the substrate
satisfying them is exhibited in §4.

\textbf{Evidence.} Three operating nodes derived from a port of the
Ordering bounded context of \texttt{dotnet/eShop} to a journaled-program
substrate, bootstrapped by paper QR codes and authorized by human
decisions, execute the same domain behavior as the original microservice
across its operational lifecycle, without a node fulfilling the
server-role at any point in time. The original deployment topology
required at least one such role; the ported topology requires none.

\subsection{Claims this paper makes}\label{claims-this-paper-makes}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The server-role in distributed software is an accidental category: it
  exists contingently, as a consequence of choices about what nodes
  replicate, rather than structurally, as a property of the problem
  being solved (§3).
\item
  The property naming this distinction --- \emph{accidental category}
  --- is operationalized by two conditions: contingency (the role's
  necessity is traceable to a specific representational choice) and
  dissolution (the role loses justification when that choice is
  replaced) (§3).
\item
  Under a class of systems in which nodes replicate programs rather than
  data, the conditions hold for the server-role: it disappears as a
  structural consequence, not as a design objective (§5).
\item
  The bootstrap of a new node --- the operation that prior literature
  treats as the residual case requiring a server --- admits a
  presentation in which the information transferred crosses by analog
  means and the authorization is exercised by humans, without any
  service running on either side (§4).
\item
  An instantiation of the conditions, constructed by porting the
  Ordering bounded context of \texttt{dotnet/eShop} to a
  journaled-program substrate and deploying it across three nodes
  bootstrapped by paper QR codes, dissolves the server-role across the
  operational lifecycle (§4, §5).
\item
  An unexpected consequence of journaled programs is that existing cloud
  and microservice ecosystems cease to be architectural dependencies and
  become reusable domain libraries within the program itself (§6).
\item
  The construct is diagnostic: it enables identifying which
  architectural categories in any given system are structural and which
  are contingent, independent of the choice to act on that information
  (§7, §8).
\end{enumerate}

\subsection{1. Introduction}\label{introduction}

This paper makes a design theory contribution in the sense of Hevner,
March, Park, and Ram (2004). It identifies and names a structural
pattern that prior literature has documented case by case without
articulating as one (the construct); derives the conditions under which
the pattern dissolves (the principles); and presents an instantiation in
which those conditions hold and the pattern is absent (the existence
proof). The genre is design science research: evidence is presented in
the form of a working artifact rather than a controlled experiment. The
contribution is conceptual; the instantiation confirms that the pattern
is contingent rather than structurally necessary, not that any specific
system is the only way to dissolve it.

Production software is built around an assumption that is rarely
articulated as such: that a \emph{server-role} exists somewhere in the
system. The role is not the physical machine, nor the deployment
topology, nor any particular cloud provider. It is the architectural
function --- accepting authoritative writes, owning coordination,
mediating between clients, providing the address against which other
parties direct their traffic. Across textbooks, reference architectures,
and tooling defaults, the role is treated as a primitive of distributed
software: present by default, named when necessary, removed only at
great cost.

Prior literature has interrogated the server's \emph{privileges} --- its
asymmetric power over data, identity, or availability --- and proposed
designs that reduce those privileges. Local-first software argues that
data should belong to its user and survive without the cloud service.
Hypercore, Dat, and related peer-to-peer systems replicate append-only
logs without a privileged coordinator. Holochain inverts the
data-centric model so that the agent's source chain and a shared
distributed hash table together replace the central server. Federated
systems pluralize the server-role across cooperating hosts. Each of
these approaches treats the server-role as a \emph{target of design} ---
something to redesign, redistribute, or replace. None of them removes
the server-role as a \emph{prerequisite for software construction}. In
each case, the role's elimination is the explicit objective of the
system; the system is built to satisfy that objective.

This paper takes a different position. Under a class of systems in which
the substrate shared between nodes is the program --- not the data, not
the state, not even the operations --- the server-role does not survive
as a necessity. It is a category whose referent dissolves. \emph{In the
framing developed here, server ceases to be a prerequisite for making
software. Bootstrap --- the last role that seemed to need a service ---
can be performed by a printed QR code.}

The paper develops three claims in cascade. First, the server-role is
identified as an \emph{accidental category}: a role that exists
contingently in distributed software, as a consequence of choices about
what nodes replicate, and that has no referent under different choices.
Second, the role's disappearance is presented not as a design goal but
as a structural consequence of those different choices: in the class of
systems where programs are journaled and replayed against shared domain
libraries, the role's elimination is what the substrate produces, not
what the engineer pursues. Third, the displacement extends beyond the
role itself: under the same conditions, existing cloud and microservice
ecosystems cease to function as architectural dependencies and become
reusable domain libraries that programs can invoke at will. The cloud is
not removed; its position in the system is restated.

The instantiation used to confirm these claims --- a port of the
Ordering bounded context from Microsoft's \texttt{dotnet/eShop}
reference application, executed across three nodes that share a
journaled domain library and bootstrap each other through paper QR codes
--- is presented as existence proof, not as the contribution of the
paper. The same construct admits other instantiations: any system that
satisfies the conditions derived in §3 would dissolve the server-role in
the same way. The choice of \texttt{dotnet/eShop} is rhetorical --- a
canonical cloud-microservices reference whose original deployment
topology makes the contrast visible --- rather than essential to the
argument. The framework underlying the instantiation was not designed to
support this paper's claim. The property is identified retrospectively,
after a chain of unrelated architectural decisions converged on a system
in which the server-role had become contingent. The work of this paper
is, in that sense, a forensic reading of a class of artifacts in which
the property happens to hold.

Section 2 situates the paper among prior work on local-first,
peer-to-peer, agent-centric, and federated systems, identifying the
dimension along which each work treats the server-role and the dimension
along which the present construct differs. Section 3 defines the
construct formally --- \emph{accidental category} --- and derives the
two conditions, contingency and dissolution, that distinguish it from a
structural requirement. Section 4 presents the instantiation: the
domain-library port, the three-node deployment, the analog bootstrap.
Section 5 examines the first consequence: the dissolution of the
server-role across the operational lifecycle. Section 6 examines the
second consequence --- the one that prior literature does not
anticipate: the conversion of existing cloud ecosystems from
architectural dependencies into reusable domain libraries. Section 7
examines limitations, counter-arguments, and the boundary of the
construct's applicability. Section 8 concludes.

\subsection{2. Related work}\label{related-work}

Local-first software (Kleppmann et al., 2019) articulates seven ideals
for software that operates primarily on devices owned by their users
rather than on remote services. Replication is mediated by Conflict-free
Replicated Data Types (CRDTs), which allow concurrent edits to merge
without coordination. The position with respect to cloud infrastructure
is precise: cloud services are admissible, but the software must
continue to function in their absence. The title of the paper ---
\emph{Local-First Software: You Own Your Data, in spite of the Cloud}
--- captures the stance: the cloud is neither rejected nor required. The
server-role is reduced to one of several optional services, and its
elimination is the explicit objective of the architectural pattern.

Peer-to-peer append-only logs constitute a second line. Secure
Scuttlebutt (Tarr, 2014) replicates per-identity signed logs across a
gossip network without a central coordinator; the log, not the data it
carries, is the unit of synchronization. Hypercore and the Dat protocol
generalize the same pattern, treating the append-only feed as a
primitive replicated by a distributed hash table. In each case the unit
of replication is a log of data --- events, document edits, file
fragments --- and the application logic that interprets the log lives in
the consuming application. The server's elimination is the design target
of the network layer.

Holochain (Harris-Braun and Brock) makes the agent the source of truth.
Each agent maintains a local source chain --- an immutable, signed
sequence of its own transactions --- and contributes to a validating
distributed hash table. The architecture inverts the data-centric
assumption of blockchain rather than addressing the cloud directly; the
elimination of the server-role is a consequence of the agent-centric
inversion. The replicated artifacts are the agent's chain and the
validated DHT; application logic, written as zomes, is interpreted
locally by each agent.

Federated systems form a fourth line. Matrix federates an event graph
across cooperating homeservers; ActivityPub federates streams of
activities; the AT Protocol of Bluesky splits hosting across a Personal
Data Server, a Relay, and an AppView, with account portability as the
displacement mechanism. The server-role is preserved but pluralized:
every participating host fulfills the role for some subset of users. The
cloud is partitioned across cooperating providers rather than
centralized or rejected.

Adjacent work admits brief mention. Solid (Berners-Lee, 2016) proposes
Personal Online Datastores that decouple identity from application
providers; the server-role moves from application owner to data host
without being eliminated. Earthstar pursues local-first synchronization
without a distributed hash table, sitting between Kleppmann's CRDT model
and Hypercore's gossip topology. Neither treats the server-role as a
category whose referent is contingent.

Each of these lines treats the server-role as a \emph{design target} ---
an architectural feature to redesign, redistribute, replace, or
pluralize. None of them names the property that the present paper
identifies: that the role is contingent on a choice about what nodes
replicate, and that under a different choice the role has no referent at
all. Table 1 summarizes the four families and the position of the
present construct.

\textbf{Table 1.} Five families of prior work and how each treats the
server-role. The bottom row identifies the construct of the present
paper; the columns describe what each family replicates between nodes,
where its application logic resides, why a privileged server is absent
(or pluralized), and how each family situates itself with respect to
cloud infrastructure.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\linewidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\linewidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\linewidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\linewidth - 8\tabcolsep) * \real{0.2000}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Family
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
What is replicated
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Where the logic lives
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
How the server-role is treated
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Relation to the cloud
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Local-first (Kleppmann) & CRDTs / data & In the application & Eliminated
as a design goal & Survival without it \\
Hypercore / Dat / SSB & Append-only logs & In the application &
Eliminated as a network goal & Replaced by peers \\
Holochain & Source chain + DHT & In the agent & Eliminated by
agent-centric inversion & Treated as an alternative architecture \\
Federated (Matrix / ActivityPub / AT Protocol) & Events / activities &
On federated servers & Pluralized across hosts & Fragmented across
providers \\
\textbf{The present construct} & \textbf{Programs} & \textbf{In the
journal of each node} & \textbf{Not a prerequisite (accidental
category)} & \textbf{Subsumed as a library} \\
\end{longtable}
}

The discriminator is the fourth column. The four families treat the
server-role as something to be designed away --- a residual feature
whose elimination is the goal. The present construct identifies the role
as contingent: under a representational choice in which nodes replicate
programs rather than data, state, or operations, the role's necessity
does not survive. The fifth column is downstream of the fourth: under
the present construct the cloud is neither evaded, replaced, nor
pluralized, but reused as a library within the program.

The substrate of §4 combines the actor model (Hewitt, 1973), event
sourcing (Fowler, 2005), and CQRS (Young, 2010), synthesized as
practical patterns by Vernon (2015). An additional move --- domain
classes carrying no persistence concerns, discovered by reflection ---
antedates the synthesis by roughly a decade in the framework underlying
the instantiation; that genealogy is described in §4.0.

This paper is the seventh in a series. Papers 1--6 develop the
substrate-level foundations on which the present construct rests,
culminating in Paper 6's \emph{infrastructural symptom}, of which the
present construct extends the discrimination from layers of
infrastructure to architectural roles.

The contribution unique to this paper is conceptual. The construct
\emph{accidental category} names a property that prior work documents
piecewise --- across local-first, peer-to-peer, agent-centric, and
federated systems --- without unifying. The two conditions, contingency
and dissolution, formalize the property. The discrimination table of §3
operationalizes it without requiring adoption of any particular
alternative. The instantiation of §4 demonstrates that the construct's
predicate dissolves under the conditions; the demonstration is not the
contribution.

\subsection{3. The construct: accidental
category}\label{the-construct-accidental-category}

A server-role is an \emph{accidental category} in a class of systems S
when both of two conditions hold.

\textbf{The contingency condition.} The role's necessity in S is
traceable to a specific representational choice --- what the nodes of S
replicate in order to share state. The choice must be identifiable: not
\emph{``the system uses a server because that is how distributed systems
work''}, but \emph{``the system uses a server because the nodes
replicate property R, and the role compensates for the coordination that
R imposes.''} When the contingency condition holds, asking \emph{what
would happen if R were replaced?} yields a non-trivial answer.

\textbf{The dissolution condition.} When that representational choice is
replaced --- by adopting a substrate in which R is replaced by a
different property --- the role loses justification. The substitution's
effect must be sufficient: nothing else about the problem domain,
workload, or operational context needs to change for the role to become
unnecessary.

Both conditions must hold. Contingency without dissolution describes a
role whose justification survives substrate change --- a structural
requirement of the problem domain. Dissolution without contingency is
not constructible: a role cannot lose justification under a substrate
change unless its justification was tied to the substrate in the first
place. \textbf{The boundary between accidental category and structural
requirement is not a property of the role-as-name but of the
function-for-which-it-is-used.}

The construct is not a universal accusation. Many roles in distributed
systems are structural requirements --- identity providers backed by
cryptographic root authority, gateways mediating external systems,
computationally expensive services such as analytics or inference
engines. No substrate change redistributes the authority of an identity
provider, removes the integration problem of an external network, or
makes intrinsically expensive computation cheap.

The discrimination operates per role-in-use-case. The same node,
identified in a deployment as ``API server,'' instantiates an accidental
category when deployed to accept authoritative writes against a shared
database, and a structural requirement when deployed to enforce
cryptographic identity claims. The discriminator is the function, not
the name.

Seven categories of architectural role are encountered regularly in
production deployments and admit discrimination under the two
conditions. The third column describes, in substrate-property terms
rather than system terms, what a substrate satisfying the two conditions
offers in place of each role.

\textbf{Table 2.} Discriminator: seven canonical categories of
architectural role under the two conditions of §3. The table identifies,
for each role, the representational choice that necessitates it and the
substrate property that would render it unnecessary.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Architectural role
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Representational choice that necessitates it
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
What a substrate satisfying the conditions offers instead
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Server as write authority & Nodes replicate data; consistency requires a
single point of acceptance & Each node accepts writes locally; the
journal records the program, which replays deterministically on every
node \\
Master / primary in replication & Nodes replicate state diffs; a primary
must order them & Each node owns its journal; cross-node causality is
recorded by reference, not orchestrated \\
Coordinator in distributed transactions & Multiple writers contend for a
logical entity across processes & Each logical entity has a single point
of authority intrinsic to the substrate; serialization is not
externalized \\
Stateful gateway & Clients cannot directly invoke deferred work without
losing causal record & Deferred work is journal-recorded; the substrate
carries causality, not the gateway \\
Application server / runtime container & The runtime, the framework, and
the application require disparate substrates & A minimal substrate
suffices to run the domain artifact \\
Bootstrap service & New nodes require an issuer of credentials and
addresses & Bootstrap is information transfer that crosses by any means,
including analog artifacts; no service runs \\
Orchestrator as coordination substrate & Stateful coordination across
instances requires external choreography & Coordination is between
actors; the substrate provides location and continuity intrinsically \\
\end{longtable}
}

The seven rows are not exhaustive; service discovery, queue-as-glue, and
configuration management admit similar analysis (§7). The construct is
diagnostic, not prescriptive: it enables auditing existing systems under
the two conditions, independently of any decision to adopt the substrate
of §4.

\subsection{4. Instantiation}\label{instantiation}

\subsubsection{4.0 Genealogy}\label{genealogy}

The substrate underlying the present instantiation is the journaled
actor-native framework whose architectural lineage is documented in
Paper 6 §4.0 and is not retraced here. The implementation source will be
released open-source with the publication of this paper; the references
in this section to specific test methods, file paths, and orchestrator
scripts are pointers into the forthcoming release. Reproduction details
are consolidated in Appendix A.

\subsubsection{4.1 The substrate}\label{the-substrate}

The substrate provides three properties that together satisfy the two
conditions of §3 for the server-role. First, each node owns a local
journal of programs --- DSL scripts that, on replay, reconstruct the
node's state deterministically. Second, the domain library --- the
classes and verbs that the journal scripts invoke --- is loaded by
reflection at startup; no node hosts the domain as a service for others.
Third, replication is journal-to-journal: nodes exchange script
references and parameters, not state diffs, and reach convergence by
replaying the same sequence of scripts against the same library. None of
these properties requires a node to fulfill the server-role; each is
intrinsic to the substrate, not externalized.

The substrate is described in the abstract throughout this section. Its
concrete instantiation as the framework that this paper's existence
proof builds upon is the one developed across the preceding papers of
the series. The choice of that framework is rhetorical --- it is the one
whose source is available and whose author can speak to its properties
--- not essential to the construct.

\subsubsection{4.2 The port}\label{the-port}

The instantiation ports the \emph{Ordering} bounded context of
Microsoft's \texttt{dotnet/eShop} reference application --- a canonical
example of microservice-oriented cloud architecture --- onto the
substrate of §4.1. The diagnostic value of the port is what it makes
visible: the lines of the original codebase that the substrate cannot
host without modification are a small minority, concentrated in
persistence-and-deployment glue. The aggregate, the value objects, the
lifecycle verbs, and the invariants move into the journaled substrate
without modification. What was previously named ``the Ordering
microservice'' turns out to be a clean domain library wrapped in
infrastructure compensating for a different substrate; the port
separates the two.

\subsubsection{4.3 The three-node
deployment}\label{the-three-node-deployment}

Three identical processes --- each running the same substrate with the
ported domain library loaded by reflection --- deploy across three
Docker containers. The coordination mesh is fully connected: three
pairwise channels forming the complete \texttt{N(N–1)/2} graph at
\texttt{N\ =\ 3}. Three is the minimum peer count at which the
dissolution claim is unambiguous: with two peers, any rotation can be
read as one serving the other; with three, no subset of two dominates
the third.

\begin{verbatim}
  ┌─────────────┐    coord pair    ┌─────────────┐
  │ ordering-a  │◄────────────────►│ ordering-b  │
  └──────┬──────┘                  └──────┬──────┘
         │ coord pair                     │
         └──────────┐         ┌───────────┘
                    ▼         ▼
                  ┌─────────────┐
                  │ ordering-c  │
                  └─────────────┘
\end{verbatim}

\textbf{Figure 1.} The three-node coordination mesh. Each node is a
peer; none is privileged.

No node carries the write-acceptance privilege permanently; the role
rotates symmetrically across peers. After the rotation completes, the
three journals are byte-coincident.

\subsubsection{4.4 The analog bootstrap}\label{the-analog-bootstrap}

The bootstrap of a new peer proceeds in five phases. \textbf{F1} opens
an invitation on the issuer side. \textbf{F2} transmits the invitation
to the joining peer by any means, including analog ones --- printed
paper, a verbal dictation, an email. \textbf{F3} opens an authenticated
connection back to the issuer. \textbf{F4} pauses for a human decision
to approve or reject. \textbf{F5} seals a credential to the joining
peer's public key and closes.

Two of the five phases are sub-software: F2 (the information transfer)
and F4 (the authorization decision). Neither requires a service to be
running anywhere; the carrier of F2 is whatever the operator chooses,
and the decision-maker of F4 is human. The three software-mediated
phases (F1, F3, F5) are local to the issuer process, which exits when
the bootstrap is complete. \textbf{The issuer process does not act as a
server, coordinator, authority, registry, or rendezvous point after
bootstrap; its only role is to transfer a credential that allows peers
to recognize each other.}

After all three containers complete the handshake --- each with its own
QR code carried by analog means and approved by a human --- the issuer
process exits. The containers continue to operate.

\begin{verbatim}
  ordering-a   Up 11 seconds   0.0.0.0:6443->6443/tcp
  ordering-b   Up 11 seconds   0.0.0.0:6444->6443/tcp
  ordering-c   Up 11 seconds   0.0.0.0:6445->6443/tcp
\end{verbatim}

\textbf{Output of \texttt{docker\ compose\ ps} after the issuer exits.}
No surviving network path connects the containers to the issuer; the
bootstrap papers are in a drawer.

\emph{In the framing developed here, the information that constituted
the bootstrap traveled by paper; the decision that authorized it was
made by a human; the cryptographic protection of the credential happened
during the brief lifetime of the issuer process. The artifact that
survives is the three running containers.}

\subsubsection{4.5 The empirical matrix}\label{the-empirical-matrix}

The demo runs across a two-by-two matrix: \emph{in-process} versus
\emph{cross-container}, and \emph{inline} versus \emph{parametric}. Each
cell is a complete execution of the workload across the three nodes; the
in-process row reduces the bootstrap to direct invocation, while the
cross-container row runs the full §4.4 protocol.

\textbf{Table 3.} Empirical matrix of the existence proof. Each cell
reports the final journal entry count, identical on all three peers and
byte-coincident on disk.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Workload regime
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
In-process (3 Stages, single OS process)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cross-container (3 Dockers, real TLS + analog bootstrap)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Inline (literal scripts) & 22 entries & 22 entries \\
Parametric (Define + bind) & 7 entries & 7 entries \\
\end{longtable}
}

The four cells produce identical final entry counts per workload regime,
and the three peers in each cell are byte-coincident on disk. The
cross-container row confirms that the operational properties of the
substrate survive the addition of the analog bootstrap and the real
cryptographic stack; the in-process row is the structural rehearsal that
isolates the substrate's intrinsic properties from the network. The
compaction from inline to parametric is real and measured; its precise
ratio and the conditions under which it varies are treated in Paper 2.

In the parametric regime, the journal records seven entries rather than
five: each rotation Director emits its own Define + Invocation pair,
because the Define cache is local to the Stage that emits it, not
replicated. The journal is the catalog of each peer's declarative
vocabulary; the cache's compactness lives on the Stage axis, not on the
journal axis.

\subsubsection{4.6 Partition tolerance and
re-join}\label{partition-tolerance-and-re-join}

The same artifact additionally exhibits partition tolerance and re-join
from disk. A peer can exit the cluster while the others continue
operating; a fresh process can mount the absent peer's data directory
and rehydrate its journal-relative state from disk alone, without
contacting the issuer or the surviving peers; and the rehydrated peer
integrates into the still-running cluster via peer-to-peer catch-up. The
five-phase decomposition of this scenario and its implications for the
construct's dissolution claim are described in §5. The relevant
observation for §4 is that the same artifact that produces the four
cells of Table 3 also produces, without modification, the operational
signature of partition tolerance and re-join from disk.

\subsection{5. The first consequence: dissolution of the
server-role}\label{the-first-consequence-dissolution-of-the-server-role}

§3 defined the server-role as a composite of four architectural
functions. §4 instantiated a substrate in which each of these functions
is distributed intrinsically rather than externalized to a privileged
node. This section reports the dissolution as it occurs across the
operational lifecycle of the cluster.

Acceptance of authoritative writes and ownership of coordination
dissolve together under Director rotation. The Director role traverses
the three peers in turn; no node carries the write-acceptance privilege
permanently; the privilege is a transient mode that any peer can assume.
The same property holds across the workload's non-happy paths ---
cancellation, error recovery --- not only across its main sequence.

Mediation between clients and provision of the cluster's address
dissolve into the peer-to-peer mesh. No node carries an authoritative
address for the cluster as a whole; a request directed at any peer is
acceptable and that peer can act as Director for it. The ``address
against which other parties direct their traffic'' --- the fourth limb
of the §3 definition --- is not a single address but the union of three
pinned addresses, none of which is privileged.

The strongest empirical signature of the role's dissolution is the
partition-tolerance scenario introduced in §4.6. One peer leaves the
cluster; the remaining two continue to operate; the absent peer rejoins
by reading its on-disk journal. Five phases describe the sequence:

\begin{itemize}
\tightlist
\item
  \textbf{R1.} The three peers converge to a shared journal state.
\item
  \textbf{R2.} One peer exits; the others continue. The cluster retains
  quorum, retains a Director, and accepts new work.
\item
  \textbf{R3.} The surviving Director issues additional work; the two
  surviving journals advance; the absent peer's on-disk journal is
  unchanged.
\item
  \textbf{R4.} A fresh process --- distinct in-memory identity, same
  data directory --- reads the disk and reconstructs the
  journal-relative state of the absent peer, without contacting the
  issuer and without contacting the surviving peers.
\item
  \textbf{R5.} The rehydrated peer integrates into the still-running
  cluster via peer-to-peer catch-up; the gap is replayed from a
  surviving peer; subsequent work brings all three peers back to a
  shared state.
\end{itemize}

The properties exercised by this scenario are not separable. The
cluster's continued operation in the absent peer's absence is only
possible if no node is the cluster's source of authority. The
rehydration without consulting the issuer is only possible if the disk
is a self-contained witness of the peer's prior participation. The
catch-up between peers is only possible if any surviving peer is capable
of replaying the gap; no specialized recovery node is invoked.
\emph{Operationally, the scenario shows that a peer can leave the
cluster, the cluster can continue accepting work in its absence, and the
peer can re-join by reading the journal that survived its disappearance
on its own disk --- no central authority is consulted at any point in
this loop. The role of the issuer was discharged once, at the original
onboarding, and is not invoked again.}

The issuer is invoked exactly once per peer. The credential sealed
during the bootstrap of §4.4 is preserved across restarts, and the
peer's continued participation is verified by the journal's
cryptographic integrity, not by reference to a running service. If a
peer loses its disk, it re-onboards as a new peer; if it preserves its
disk, it returns as itself, regardless of how long it was absent.

The four functions of the §3 definition, taken together, are what the
literature names ``server.'' The §4 substrate dissolves them in turn:
write acceptance and coordination ownership distribute across the three
peers via rotation; mediation and addressing distribute across the mesh;
partition tolerance and re-join distribute across the surviving peers
via catch-up; the bootstrap issuer participates once and is absent
thereafter. \textbf{The server is not a permanent identity. The server
is a transferable coordinator role that any peer can assume in turn ---
and any peer can leave the cluster and re-join without reference to a
coordinator outside it.}

\subsection{6. The second consequence: cloud as library, not
infrastructure}\label{the-second-consequence-cloud-as-library-not-infrastructure}

A second operational consequence becomes visible in §4 once the cluster
continues to operate after bootstrap without any of the services that
originally constituted the Ordering microservice. Cloud and microservice
ecosystems, when accessed from a journaled program, are observable as
libraries invoked by the program rather than as services required for
the cluster to exist.

The difference becomes operationally visible in what the cluster does
when the external system is unreachable. A dependency's absence is an
outage: the dependent system stops functioning because its precondition
is missing. A library's absence is a failed invocation: the program
records the failure and continues. The two relations differ in what the
dependent system does at the moment of unavailability.

After the port described in §4.2, none of the services that previously
had to be running for the Ordering microservice to exist are reachable
from the cluster. The domain code runs entirely inside each node; what
previously required an API host, a database, and an event bus is now a
library invoked by journaled programs. The cluster operates against its
own journals; the cloud topology of the original microservice has no
point of contact with the cluster after the bootstrap of §4.4 completes.

The same relation generalizes to external services that the cluster
invokes during normal operation. If the identity provider is unreachable
at the moment of invocation, the journal records the failed invocation
and the cluster continues. If a payment gateway is unreachable, the
journal records the attempt and its outcome. The external service's
absence is recorded as data, not experienced as outage. The cluster does
not inhabit any of these services; it invokes them at chosen moments
through journaled Reactions (Paper 3) and proceeds with their result,
including the result of their absence.

This operational relation inverts the formulation that has organized the
local-first software literature:

\textbf{Table 4.} The inversion of Kleppmann's formulation, captured as
a six-word pair.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.5000}}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.5000}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Kleppmann (Local-first)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
The present construct
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\emph{Own your data in spite of the cloud.} & \emph{Use the cloud
without depending on it.} \\
\end{longtable}
}

Local-first software treats the cloud as something the software must
survive without. The cluster of §4 neither survives without the cloud
nor depends on it; it invokes cloud services when useful and proceeds
without them when not. The two relations differ in what the absence of a
cloud service means for the cluster's continued existence.

The experiment shows that cloud services, when used from a journaled
program, are consumed as invocable capabilities rather than as the
substrate in which the program must reside.

\subsection{7. Limitations and
counter-arguments}\label{limitations-and-counter-arguments}

The instantiation of §4 carries specific limitations that follow from
the choices made for the demo. They are reported here honestly; none of
them affects the construct's predicate, but each affects what the
present experiment alone establishes.

\textbf{Identity in the port is local.} The original
\texttt{dotnet/eShop} deployment uses an external identity provider; the
port substitutes a local string-typed identifier. The construct does not
predict that identity providers dissolve under the conditions of §3 ---
identity providers backed by cryptographic root authority are structural
requirements, as Table 2 notes --- and the demo does not exhibit one. A
second experiment that integrates an external identity provider as a
Reaction-invoked library would extend the empirical surface without
changing the construct's claim.

\textbf{Cross-context dependencies are not portrayed.} The Ordering
bounded context of \texttt{dotnet/eShop} carries references resolved by
the Catalog and Basket bounded contexts; the port treats those
references as opaque identifiers. The construct predicts that the same
port mechanism applies to Catalog and Basket; the present experiment
exhibits Ordering only.

\textbf{The empirical matrix is at N = 3.} The mesh in §4.3 connects
three peers --- the minimum non-trivial fully-connected mesh. The
construct's predicate is not bounded at N = 3; it dissolves the
server-role whenever the conditions of §3 hold for the substrate, which
is independent of node count. Scaling to larger N changes the network
cost of the mesh; it does not change the conditions of §3.

\textbf{The cross-container partition-tolerance scenario is captured in
a parallel artefact, not in the matrix of §4.} The natural
cross-container equivalent of the in-process scenario in §4.6 ---
stopping a container, allowing the running cluster to continue accepting
work in its absence, restarting the container, and watching it rehydrate
from its mounted volume and rejoin via catch-up against a surviving peer
--- is implemented in the same orchestrator that produced the matrix
(the \texttt{-\/-rehydrate-demo} flag of the demo script) and depends on
no substrate feature that is not already exercised by the in-process
test in §4.6. The decision to keep §4 to the four cells of the matrix,
rather than expanding to a fifth cell, was rhetorical: the in-process
exercise establishes the property structurally, and §5 reasons from
there. The cross-container capture documents the same property under a
different process boundary; it does not add a new claim.

\textbf{The compaction ratio between inline and parametric workloads
depends on Director rotation.} Each new Director journals a fresh Define
+ Invocation pair, so the Define cache compacts on the Stage axis rather
than on the journal axis. The compaction is real and measured at three
rounds; its precise value at larger workloads is treated in Paper 2.

A reviewer may object: \emph{did the experiment really remove the
server, or did it rewrite the controllers under a different name?} The
discriminator is operational. The server-role of §3 is measured by what
happens when the candidate node is removed. After the bootstrap of §4.4,
the issuer process exits; no node in the running cluster fulfills the
four functions of the §3 definition; the partition-tolerance scenario of
§4.6 shows that the cluster continues to operate when any single peer is
removed. The role is not present in the topology to be removed; it does
not have an incumbent. The objection presupposes a role that the
experiment shows has no referent.

A second objection: \emph{identity, authority, and payment are
intrinsically central, so the cluster is not really server-free.} The
construct does not claim otherwise. §3 separates structural requirements
(identity providers, external systems, expensive computation) from
accidental categories. The cluster's discharge of the server-role for
the Ordering use-case does not imply that no node in any larger
deployment ever fulfills a server-role; it implies that the role is not
a structural property of Ordering. The same per-use-case discrimination
applies to identity: when a deployment attributes credential validation
to a cryptographic root authority, that authority's role is structural;
that fact does not propagate to the Ordering domain.

The construct's predicate extends to other architectural roles that
admit the same analysis. Service discovery, queue-as-glue between
bounded contexts, configuration-management roles, and
distributed-tracing collectors are auditable under the two conditions of
§3; the present experiment does not exercise them, and the substrate's
predicted relation to each is left as a forward question. The construct
admits this open extension by design --- its diagnostic value is in
identifying which categories in any given system are contingent, not in
enumerating them exhaustively.

\subsection{8. Conclusion}\label{conclusion}

This paper has named a property that was previously implicit in
distributed-systems practice: that the server-role is not necessarily
structural to the problem domain but can arise from representational
choices in the persistence model. The term \emph{accidental category}
designates this condition.

Two conditions have been stated that allow practitioners to distinguish
accidental categories from structural requirements, and §3 has
operationalized these conditions across canonical architectural roles.
The instantiation of §4 demonstrates that, under a journaled-program
substrate, the Ordering bounded context of \texttt{dotnet/eShop}
continues to operate across three mutually bootstrapping nodes without
any participant fulfilling the server-role. The demonstration does not
argue uniqueness of the substrate; it establishes observability of the
condition.

Under these conditions, the four functions traditionally attributed to
the server-role --- authoritative writes, coordination ownership, client
mediation, and address provision --- are observed to dissolve into
properties of the substrate itself. What appears as ``server behavior''
is redistributed across peer rotation, mesh communication, disk-backed
partition tolerance, and one-time bootstrap discharge.

The same observation extends to external infrastructure. Cloud services,
microservices, and infrastructural ecosystems, when accessed from a
journaled program, are no longer required habitats for the system to
exist, but libraries invoked by the program when needed. The distinction
is not eliminative but relational: from inhabitation to invocation.

The contribution of this paper is the formalization of a discrimination
that can be applied independently of the substrate used in §4. The
construct allows any production system to be examined role-by-role,
asking whether each role is tied to the problem being solved or to the
persistence model chosen.

The discrimination is per role-in-use-case, not per system. Identity
authorities, payment networks, sensor integrations, and computationally
intensive services remain structural under any substrate. The construct
predicts no universal dissolution; it predicts examinability.

Paper 6 applied the construct at the layer level, identifying many
infrastructural layers --- caches, queues, locks, application servers,
orchestration clusters --- as compensations for the underlying
persistence model. The present paper extends the same discrimination to
the architectural role that made such layers necessary in the first
place: the server. The two papers apply one construct at two levels of
abstraction --- first to layers, then to roles. From this point on,
treating the server-role as a structural requirement is no longer an
assumption but a choice that can be made explicit and evaluated.

\subsection{Appendix A ---
Reproducibility}\label{appendix-a-reproducibility}

The artifact described in §4 will be released open-source with the
publication of this paper. The references in this appendix to file
paths, test method names, and orchestrator scripts are pointers into the
forthcoming release. The artifact comprises the ported domain library,
the host process, the orchestrator script, the test suite, and the
supporting documentation in \texttt{notes/}.

\subsubsection{A.1 Source and suite}\label{a.1-source-and-suite}

The release will include the substrate framework, the ported
\texttt{Ordering} domain library, the demo orchestrator, the test suite,
and the supporting documentation referenced in this appendix. The exact
branch, commit, and test-count metadata will be recorded with the
release tag at the time of publication.

\subsubsection{A.2 Demo orchestrator}\label{a.2-demo-orchestrator}

The orchestrator script \texttt{docker/run-demo.sh} provisions three
Docker containers, runs the analog bootstrap handshake against each, and
exercises one cell of the matrix per invocation. Three flags select the
scenario:

\begin{itemize}
\tightlist
\item
  \texttt{bash\ docker/run-demo.sh} --- cross-container inline (22
  entries final).
\item
  \texttt{bash\ docker/run-demo.sh\ -\/-parametric} --- cross-container
  parametric (7 entries final).
\item
  \texttt{bash\ docker/run-demo.sh\ -\/-rehydrate-demo} --- captures the
  cross-container leave-and-rejoin scenario referenced in §7: stops one
  container, allows the cluster to advance, restarts the container, and
  observes its rehydration and catch-up.
\end{itemize}

Each invocation runs in approximately 60--90 seconds on Docker Desktop.

\subsubsection{A.3 Tests cited}\label{a.3-tests-cited}

The empirical claims in the body are exercised by specific tests in the
suite:

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.2500}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Body section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Property
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Test method
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Project
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
§4.4 & F1--F5 bootstrap handshake under real cryptography &
\texttt{EndToEnd\_RealCryptoOverRealTls\_RoundsTripIdentity} &
\texttt{UnitTestChoreography/UsherOnboardingTests.cs} \\
§5 (single peer operates alone) & Force-promoted Stage operates without
peers & \texttt{SingleStage\_OperatesAlone\_WhenPromotedForce} &
\texttt{UnitTestPaper7EShop/ThreeNodeOrderingTests.cs} \\
§4.5 (in-process, inline) & Three-stage Director rotation, inline regime
& \texttt{ThreeStages\_DirectorRotation\_AllConverge\_HappyPath} &
\texttt{UnitTestPaper7EShop/ThreeNodeOrderingTests.cs} \\
§4.5 (in-process, parametric) & Three-stage Director rotation,
parametric regime &
\texttt{ThreeStages\_DirectorRotation\_AllConverge\_HappyPath\_Parametric}
& \texttt{UnitTestPaper7EShop/ThreeNodeOrderingTests.cs} \\
§5 (rotation under cancellation) & Cancellation branch replicates across
three stages &
\texttt{CancellationBranch\_Replicates\_AcrossThreeStages} &
\texttt{UnitTestPaper7EShop/ThreeNodeOrderingTests.cs} \\
§4.5 (instrumentation) & DLL load + per-phase convergence + journal
bytes snapshot & \texttt{Metrics\_F1\_Snapshot} &
\texttt{UnitTestPaper7EShop/ThreeNodeOrderingTests.cs} \\
§4.6 / §5 (partition tolerance) & Leave-and-rejoin from disk,
peer-to-peer catch-up &
\texttt{OneStage\_Offline\_OthersAdvance\_Rehydrate\_CatchUp\_Parametric}
& \texttt{UnitTestPaper7EShop/ThreeNodeOrderingTests.cs} \\
§5 (rehydration without onboarding) & Rehydrated Stage restores identity
from disk alone &
\texttt{RehydratedStage\_RestoresIdentityAndJournal\_WithoutOnboarding}
& \texttt{UnitTestPaper7EShop/ThreeNodeOrderingTests.cs} \\
\end{longtable}
}

\subsubsection{A.4 Cryptographic stack}\label{a.4-cryptographic-stack}

The cryptographic primitives underlying the F1--F5 handshake of §4.4 are
documented in \texttt{Choreography/Usher/Crypto/*.cs}. The
implementations use BouncyCastle 2.6.2:

\begin{itemize}
\tightlist
\item
  Ed25519 keypair generation, signing, verification:
  \texttt{Ed25519StageKeyGenerator.cs}, \texttt{Ed25519Signer.cs},
  \texttt{Ed25519SignatureVerifier.cs}.
\item
  Ed25519-to-X25519 derivation (Edwards-to-Montgomery, \emph{u = (1 + y)
  / (1 − y) mod p}): \texttt{Ed25519ToX25519.cs}.
\item
  Sealed-box AEAD using X25519 ECDH and ChaCha20-Poly1305, with the
  participating public keys bound as additional authenticated data:
  \texttt{SealedBoxPayloadSealer.cs}.
\item
  Kestrel TLS listener and self-signed certificate generation with SAN:
  \texttt{HttpsTransportListener.cs}, \texttt{SelfSignedCert.cs}.
\item
  TOFU certificate fingerprint pin via
  \texttt{SocketsHttpHandler.SslOptions.RemoteCertificateValidationCallback}:
  \texttt{HttpsClientFactory.cs}.
\end{itemize}

\subsubsection{A.5 Share-link URI}\label{a.5-share-link-uri}

The share-link emitted by the issuer takes the form
\texttt{puppeteer-usher://v1/\{base64url(json)\}}. The JSON payload
carries the transport address of the issuer endpoint and a SHA-256
fingerprint (\texttt{fp} field) of its TLS certificate. The
encoder/decoder is defined by \texttt{IUsherShareLinkEncoder} in
\texttt{Choreography/Usher/}. A representative share-link under load is
449 characters in length, within the alphanumeric capacity of a QR code
at version 14 with error correction level M (approximately 535
characters).

\subsubsection{A.6 Wire format}\label{a.6-wire-format}

The HTTPS transport between containers carries a \texttt{ChannelPurpose}
byte after the sender identifier in each frame. This wire format (v2)
prevents Define and Invocation channels from collapsing into the same
channel-dictionary key when Director rotation runs the parametric
workload across multiple containers; the in-process regime is
unaffected. The wire format is implemented in the transport handlers
under \texttt{Choreography/Transport/Https/}.

\subsubsection{A.7 Supporting reports}\label{a.7-supporting-reports}

Two markdown reports in the \texttt{notes/} directory of the same branch
document the experiment in detail:

\begin{itemize}
\tightlist
\item
  \texttt{notes/paper7\_phase1\_results.md} --- the in-process row of
  the matrix (eleven sections plus three addenda).
\item
  \texttt{notes/paper7\_phase2\_results.md} --- the cross-container row,
  the hardening cycle that converged on the demo, the three-Docker mesh,
  the rotation protocol, and the parametric cell.
\end{itemize}

\subsubsection{A.8 Recordings}\label{a.8-recordings}

Eleven cast recordings (asciinema format) and eleven GIF renders capture
the four cells of the matrix of §4.5 and the rehydration property of
§4.6, one recording per Docker container per cell. The recordings are
published alongside this paper as supplementary material, hosted in the
\texttt{paper7-assets/} directory of the \texttt{puppeteer-papers}
repository.

\subsection{Code provenance}\label{code-provenance}

Source-code references in this paper resolve against the public
Puppeteer repository at commit
\href{https://github.com/alvaroNCubo/puppeteer/tree/2f31f9674a5de816bdf1bf9d8360ff218a02e4da}{\texttt{2f31f96}}
(2026-05-18). The snapshot is archived in Software Heritage under the
following persistent identifier:

\begin{verbatim}
swh:1:dir:10e7e6bad7eb77b6c2e406762026177f95c3ae92;
  origin=https://github.com/alvaroNCubo/puppeteer;
  anchor=swh:1:rev:2f31f9674a5de816bdf1bf9d8360ff218a02e4da
\end{verbatim}

Inline references of the form \texttt{file.cs:NN} (e.g.,
\texttt{ActorHandler.cs:38}) resolve against this snapshot. A reader can
construct a per-file SWHID by adding the qualifiers
\texttt{;path=\textless{}path\textgreater{};lines=\textless{}NN\textgreater{}}
to the directory SWHID above. Future commits to the repository may
renumber lines; the SWHID preserves the cited state independently of any
future change to the repository or its hosting.

\subsection{Acknowledgments}\label{acknowledgments}

The author used large language models (including Claude and ChatGPT) as
editorial assistants for language refinement, structural feedback, and
literature navigation. All original ideas, terminology, theoretical
constructs, and technical content presented in this work are solely the
author's.

\subsection{References}\label{references}

Fowler, M. (2005). Event sourcing. Retrieved from
https://martinfowler.com/eaaDev/EventSourcing.html

Harris-Braun, E., Luck, N., \& Brock, A. (2018). \emph{Holochain:
Scalable agent-centric distributed computing} (Alpha 1 white paper).
Holo.
https://www.holochain.org/documents/holochain-white-paper-alpha.pdf

Hevner, A. R., March, S. T., Park, J., \& Ram, S. (2004). Design science
in information systems research. \emph{MIS Quarterly}, 28(1), 75--105.

Hewitt, C., Bishop, P., \& Steiger, R. (1973). A universal modular ACTOR
formalism for artificial intelligence. \emph{Proceedings of the 3rd
International Joint Conference on Artificial Intelligence (IJCAI)},
235--245.

Kleppmann, M., Frazee, P., Gold, J., Graber, J., Holmgren, D., Ivy, D.,
Johnson, J., Newbold, B., \& Volpert, J. (2024). Bluesky and the AT
Protocol: Usable decentralized social media. In \emph{Proceedings of the
ACM ConEXT-2024 Workshop on the Decentralization of the Internet}. ACM.
https://doi.org/10.1145/3694809.3700740

Kleppmann, M., Wiggins, A., van Hardenberg, P., \& McGranaghan, M.
(2019). Local-first software: You own your data, in spite of the cloud.
In \emph{Onward! 2019: Proceedings of the 2019 ACM SIGPLAN International
Symposium on New Ideas, New Paradigms, and Reflections on Programming
and Software} (pp.~154--178). ACM.
https://doi.org/10.1145/3359591.3359737

Lemmer-Webber, C., Tallon, J., Shepherd, E., Guy, A., \& Prodromou, E.
(2018). \emph{ActivityPub} (W3C Recommendation, 23 January 2018). World
Wide Web Consortium. https://www.w3.org/TR/activitypub/

Matrix.org Foundation. (n.d.). \emph{Matrix specification} {[}Online
document{]}. Retrieved from https://spec.matrix.org/

Ogden, M., McKelvey, K., \& Madsen, M. B. (2017). \emph{Dat: Distributed
dataset synchronization and versioning} {[}White paper{]}. Code for
Science. https://github.com/datprotocol/whitepaper

Rivera, A. (2026a). Anti-porous architecture: a unified design principle
for CQRS + Actor + Event-Sourcing systems. \emph{Puppeteer Papers
Series}, Paper 1.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/01-anti-porosity.md

Rivera, A. (2026b). Program--value separability: the structural
precondition for compilation, caching, and dense journaling in a DSL
runtime. \emph{Puppeteer Papers Series}, Paper 2.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/02-program-value-separability.md

Rivera, A. (2026c). Reactions and the partition: opt-in eventual
consistency in actor-native systems. \emph{Puppeteer Papers Series},
Paper 3.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/03-reactions-and-partition.md

Rivera, A. (2026d). Preserving semantic continuity across actors: a
tell-based approach without orchestration. \emph{Puppeteer Papers
Series}, Paper 4.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/04-cross-actor-continuity.md

Rivera, A. (2026e). The journal as substrate: unifying deployment,
replication, backup, and offline operation in distributed systems.
\emph{Puppeteer Papers Series}, Paper 5.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/05-substrate-operations.md

Rivera, A. (2026f). Most infrastructure layers are symptoms of the
persistence model: a construct for auditing production stacks.
\emph{Puppeteer Papers Series}, Paper 6.
https://github.com/alvaroNCubo/puppeteer-papers/blob/main/06-infrastructural-symptom.md

Sambra, A. V., Mansour, E., Hawke, S., Zereba, M., Greco, N., Ghanem,
A., Zagidulin, D., Aboulnaga, A., \& Berners-Lee, T. (2016).
\emph{Solid: A platform for decentralized social applications based on
linked data} (Technical Report). MIT CSAIL \& Qatar Computing Research
Institute. http://emansour.com/research/lusail/solid\_protocols.pdf

Tarr, D., Lavoie, E., Meyer, A., \& Tschudin, C. (2019). Secure
Scuttlebutt: An identity-centric protocol for subjective and
decentralized applications. In \emph{Proceedings of the 6th ACM
Conference on Information-Centric Networking (ICN '19)} (pp.~1--11).
ACM. https://doi.org/10.1145/3357150.3357396

Vernon, V. (2015). \emph{Reactive messaging patterns with the actor
model: Applications and integration in Scala and Akka}. Addison-Wesley.

Young, G. (2010). \emph{CQRS documents}. Retrieved from
https://cqrs.files.wordpress.com/2010/11/cqrs\_documents.pdf

\end{document}
